---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# climecol

<!-- badges: start -->
[![R-CMD-check](https://github.com/jbaafi/climecol/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/jbaafi/climecol/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/jbaafi/climecol/graph/badge.svg)](https://app.codecov.io/gh/jbaafi/climecol)
<!-- badges: end -->

## Overview

`climecol` provides tools for archiving, cleaning, analyzing, and visualizing weather and climate data for ecological and infectious-disease modeling.
It ships curated daily weather data and simple helpers for quick plotting and analysis.

## Installation

You can install the development version of `climecol` from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("jbaafi/climecol")
```

(Alternatively: `pak::pak("jbaafi/climecol")`.)

## Example

Load the package and explore the included Newfoundland dataset (2008–2023):

```{r example}
library(climecol)

# Load dataset
data(weather_nl)

# Peek at structure
dplyr::glimpse(weather_nl)

```

Plot daily rainfall:

```{r rainfall-plot, message=FALSE, warning=FALSE}
plot_rainfall(weather_nl)
```

> This figure was generated when knitting this README.  

## Contributing

Issues and pull requests are welcome via the repo’s [issue tracker](https://github.com/jbaafi/climecol/issues).

## Contact

If you’re interested in discussing collaborations or have questions about `climecol`, email me at [jbaafi@mun.ca](mailto:jbaafi@mun.ca).

## License

MIT © Joseph Baafi

